---
title: "Assignment 3: Data Exploration"
author: "Josh Salzberg"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup + data upload,  message = FALSE}
# for this assignment, we're using tidyverse, lubridate and here
library(tidyverse)
library(lubridate)
library(here)

# we are also using 2 datasets: ECOTOX neonicotinoid dataset & Niwot Ridge NEON dataset
ecotoxdf <- read.csv(here('./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'),
                          stringsAsFactors = T)
neondf <- read.csv(here('./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
                   stringsAsFactors = T)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We might be interested in seeing at what level of accumulations are the chemicals in insectisides effective, how the chemicals travel through food chains, or capturing a more complete view of how the chemicals enter the environment beyond the site of application.  

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We might be interested in identifying optimal component admixtures for rich decomposition. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. An elevated and ground-level pair of traps are deployed for every 400 square meter plot areas, resulting in 1-4 trap pairs per plot. 
 2. The frequency at which traps are checked is dependent on the predominant tree type at the sampling location, with deciduous sites receiving biweekly sampling, while evergreen sites are checked every 1-2 months
 3. Trap placement is either ordered or random depending on the vegetation. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5 - Finding the dimensions of Neonics dataset}
# dim() returns the overall dimensions of a dataframe
dim(ecotoxdf)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6 - summary of Effect}
#sorting the summary count of factors in the Effect column of ecotoxdf
sort(summary(ecotoxdf$Effect),decreasing = T)


```

> Answer: The 5 most common effects studied are population, mortality, behavior, feeding behavior, and reproduction. These are vital measurements to understand how an insecticide impacts insect populations. Population, mortality and reproduction are often taken together as a trio to understand demographic changes. Behavior and feeding behavior are important to capture how insecticides disrupt the natural function of the insects.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7 - Summary of species name}
# sorting the summary count of species common names
sort(summary(ecotoxdf$Species.Common.Name), decreasing = TRUE)

```

> Answer: the six most common species listed are p"Other",] Honey Bee, Parastic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee and Italian Honeybee. Ignoring "Other", we can easily identify these are all pollinators, who play a very important role in producing healthy agricultural outputs, by fertilizing the flowering plants to produce fruit. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
levels(ecotoxdf$Conc.1..Author.)
class(ecotoxdf$Conc.1..Author.)

```

> Answer: The class is currently a factor - this is because there are an apparent variety of conventions in the data. There are a mix of integers and floating points, there are slashes appended to the end of many entries (seemingly to indicate "per" the units found in the next column), and there are even inequalities and approximations of the data. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}
ggplot(ecotoxdf) +
  geom_freqpoly(aes(x = Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}
ggplot(ecotoxdf) +
  geom_freqpoly(aes(x = Publication.Year, colour = Test.Location))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The two most common test locations are Lab & "Field natural"; field natural is the predominant sourcefrom the early 1990s to 2000, when Lab becomes the predominant source for most of the remainder of the period. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
ggplot(ecotoxdf)+
  geom_bar(aes(x = Endpoint))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: NOEL No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test (NOEAL/NOEC) 
>& LOEL Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different (as reported by authors) from  responses of controls (LOEAL/LOEC)


## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}
library(lubridate)
class(neondf$collectDate)
neondf$collectDate <- ymd(neondf$collectDate)
class(neondf$collectDate)
unique(neondf$collectDate)

```

13. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}
unique(neondf$plotID)
summary(neondf$plotID)

```

> Answer: Unique provides only the list of unique values, while summary also provides counts of those unique values in the dataset

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}
ggplot(neondf)+
  geom_bar(aes(x=functionalGroup))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}
ggplot(neondf)+
  geom_boxplot(aes(x=dryMass, y=functionalGroup))

ggplot(neondf)+
  geom_violin(aes(x=dryMass, y=functionalGroup),
              draw_quantiles = c(0.25, 0.5, 0.75))



```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: the boxpolot is more effective because there don't seem to be enough observations in the dataset for the violin plot to adjust shape to frequency.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Seeds, leaves and flowers
