---
title: "Assignment 10: Data Scraping"
author: "Josh Salzberg"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse); library(rvest); library(here)
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
MyURL <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024'
MySite <- read_html(MyURL)
MySite

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
SystemName <- MySite %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
SystemName

PWSID <- MySite %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5) ") %>% 
  html_text()
PWSID

Ownership <- MySite %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

MaxDailyUse <- MySite %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
MaxDailyUse <- as.factor(MaxDailyUse)


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
MyFrame <- data.frame(Date = my(c("1, 2024","2, 2024","3, 2024", "4, 2024", 
                                  "5, 2024", "6, 2024", "7, 2024", "8, 2024",
                                  "9, 2024", "10, 2024", "11, 2024", "12, 2024")),
                      Ownership = rep(Ownership,12),
                      PWSID = rep(PWSID, 12),
                      System = rep(SystemName, 12),
                      MaxDailyUse = as.numeric(c(MaxDailyUse[[1]],MaxDailyUse[[4]],MaxDailyUse[[7]], MaxDailyUse[[10]], 
                                      MaxDailyUse[[2]],MaxDailyUse[[5]],MaxDailyUse[[8]],MaxDailyUse[[11]],
                                      MaxDailyUse[[3]],MaxDailyUse[[6]],MaxDailyUse[[9]],MaxDailyUse[[12]])))
MaxDailyUse[[1]]

#5 
MyLinePlot <- ggplot()+
  geom_line(data=MyFrame, aes(x=Date,y=MaxDailyUse))

MyLinePlot
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6.  
# Note: I am building from the structure demonstrated in the lesson file.

HarvestSite <- function(the_year, the_facility){
  
  #Retrieve the website contents 
  FunctionSite <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                                  the_facility, '&year=', the_year))
  
  #Scrape the data items
  SystemName <- FunctionSite %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  PWSID <- FunctionSite %>% html_nodes("td tr:nth-child(1) td:nth-child(5) ") %>% html_text()
  Ownership <- FunctionSite %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  MaxDailyUse <- FunctionSite %>% html_nodes("th~ td+ td") %>% html_text()
  #Convert to a dataframe
  Withdrawals <- data.frame("Month" = c(1,7,2,8,3,9,4,10,5,11,6,12),
                               "Year" = rep(the_year,12),
                               'Ownership' = rep(Ownership,12),
                                  'PWSID' = rep(PWSID, 12),
                                  'System' = rep(SystemName, 12),
                                  'MaxDailyUse' = as.numeric(c(MaxDailyUse[[1]],MaxDailyUse[[4]],MaxDailyUse[[7]], MaxDailyUse[[10]], 
                                      MaxDailyUse[[2]],MaxDailyUse[[5]],MaxDailyUse[[8]],MaxDailyUse[[11]],
                                      MaxDailyUse[[3]],MaxDailyUse[[6]],MaxDailyUse[[9]],MaxDailyUse[[12]]))) %>% 
      mutate(Date = my(paste(Month,"-", Year))) %>% 
      select(Date,Ownership, PWSID, System, MaxDailyUse)
  
  #Pause for a moment - scraping etiquette
  #Sys.sleep(1) #uncomment this if you are doing bulk scraping!
  
  #Return the dataframe
  return(Withdrawals)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
DurhamWithdrawals <- HarvestSite(2020,'03-32-010')

str(DurhamWithdrawals)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
CaryWithdrawals <- HarvestSite(2020, "03-92-020")

CombinedData <- full_join(x = CaryWithdrawals, y = DurhamWithdrawals)
CombinedData$System <- as.factor(CombinedData$System)
str(CombinedData)

ggplot()+
  geom_line(data=CombinedData, aes(x=Date, y=MaxDailyUse, color= System))


```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
Years <- c(2018:2023)
Cary6times <- rep("03-92-020",6)
CaryYears <- map2(Years, Cary6times, HarvestSite) 
LongCary <- bind_rows(CaryYears)

ggplot(data=LongCary,aes(x=Date,y=MaxDailyUse))+
  geom_line()+
  geom_smooth(method='loess')


```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Relatively steady trend, with minor increase.
>
